---
title: "ADA Final exam"
output: html_document
---

# Exercise 1.
In this exercise, we are going to evaluate whether streptomycin improves sore throat using data from a randomized controlled trial that was published in 1948. The original paper can be accessed at https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2091872/?page=1. Please run the code below to get started.
```{r}
install.packages("medicaldata")
library(medicaldata)
library(odds.n.ends)
library("MASS")
library("nnet")
library(multcomp)

#assign strep dataset in the medical data package to the environment
strep_tb<-strep_tb #See https://higgi13425.github.io/medicaldata/
head(strep_tb)

#label radnum variable that measures improvement on a six point scale with Death being 1 and Considerable improvement being 6
rad_numLabels <- c("Death","Considerable deterioration","Moderate or slight deterioration","No material change","Moderate or slight improvement","Considerable improvement")

rad_numValues <- 1:6
strep_tb$rad_numLabels <- rad_numLabels[match(strep_tb$rad_num,rad_numValues)]
strep_tb$rad_numLabels <- factor(strep_tb$rad_numLabels,levels=rad_numLabels)
table(strep_tb$rad_numLabels)

#recode control as 0 and Streptomycin	 as 1
strep_tb$arm<-ifelse(strep_tb$arm=="Control", 0, 1)
strep_tb$arm<-factor(strep_tb$arm, levels=c(0:1), labels=c("Control", "Streptomycin"))
table(strep_tb$arm)
```

# 9. How many males and females are in the strep_tb dataset?
```{r}
table(strep_tb$gender)
# There are 59 females and 48 males
```

# 10. Choose an appropriate model to model the association between arm (streptomycin vs. control) and improved (improvement vs. no improvement). a.Report the measure of association value and the 95% CI. b. Interpret the results.
```{r}
# Appropriate model is Logistic regression because the outcome is binary
model_q10<-glm(improved ~ arm, data=strep_tb, family="binomial")
summary(model_q10)
odds.n.ends(model_q10)

```

# 11. Choose an appropriate model to model the association between arm (streptomycin vs. control) and rad_num. a. Report the measure of association and value(s) and the 95% CIs. b. Interpret the results.

```{r}
strep_tb$rad_num <- as.factor(strep_tb$rad_num)
model_q11 <- polr(rad_num ~ arm, Hess=TRUE, data=strep_tb)
summary (model_q11)
tidy(model_q11, conf.int=TRUE, exponentiate = TRUE, p.values=TRUE)

# The answer will be OR 5.43 and CI 4.5 - 2.64
#The odd of considerable improvement is 5.43 times higher than the odds of death to considerable deterioration among individuals with streptomycin treatment than those without streptomycin (OR=5.43,95% CI: 2.64-11.53)
```

# Exercise 2.
This exercise will use the Framingham data to model the association between current smoking and death. The data dictionary can be found here: https://biolincc.nhlbi.nih.gov/media/teachingstudies/FHS_Teaching_Longitudinal_Data_Documentation_2021a.pdf?link_time=2021-12-02_12:02:38.620894. Please run the code below to get started.

```{r}
library(readr)
library(survminer) 
library(survival) 
library(ggfortify) 
library(lmtest)
#import the csv file
frm<-read_csv("frmgham2.csv")

frm$cursmoke<-factor(frm$CURSMOKE, levels= c(0:1), labels = c("Not Current Smoker", "Smoker"))
frm$sex<-factor(frm$SEX, levels=c(1:2), labels = c("Male", "Female"))
```

# 12. Determine number of individuals for variables needed for this analysis (cursmoke, sex, DEATH). Report the number of subjects who are a. current smokers, b. male, and c. who have died.
```{r}
# current smokers
tabyl(frm$cursmoke)
# The number of current smokers is 1705

# males
tabyl(frm$sex)
# The number of males is 1944

# Dead
tabyl(frm$DEATH)
#The number of deaths is 1550
```

# 13. Determine the mean age and TIMEDTH. Report these means (a. age, b. TIMEDTH).
```{r}
# mean age
mean(frm$AGE)
# The mean age is 59.67 years
# Mean time to death
mean(frm$TIMEDTH)
# The TIMEDTH is 7505.63

```

# 14. Generate a Kaplan Meier plot that shows survival probabilities for death by cursmoke. Upload a snapshot or picture of your plot.
- First we need to generate the survival probabilities at each time t in the datafram
```{r}
smoking.surv <- survfit(Surv(TIMEDTH, DEATH) ~ cursmoke, frm) 

summary(smoking.surv)

```

```{r}
#class(smoking.surv)
autoplot(smoking.surv) + labs(x="Days", y="Proportion Surviving", title="KM survival plots for current smoking")
```
```{r}
ggsurvplot(smoking.surv, data = frm,  pval= TRUE, conf.int=TRUE,  ggtheme = theme_minimal())
```

# 15. Test for statistical differences between survival curves using the log rank test. Report and interpret the results.

## Calculate exact log rank in R usign the survdiff function
```{r}
survdiff(Surv(TIMEDTH, DEATH) ~ cursmoke, data=frm)

# Chi square = 14.5, p-value = 0.0004
# Interpretation: There is a significant difference between the death rates between current smokers and non-current smokers (p-value<0.05).
```

# 16. Run a univariate Cox PH regression model to determine the hazard rate of DEATH in current smokers vs. non-current smokers. a. Report the HR and 95% CI and b. Interpret the results.
```{r}

cox_mod1<-coxph(Surv(TIMEDTH, DEATH)~cursmoke, frm, ties="efron") 

summary(cox_mod1)


# HR = 1.22 (95% CI 1.10 - 1.35)
# Interpretation: Respondents who were current smokers had a 1.22 (95% CI 1.1-1.35) times higher hazard of death than non-current smokers.

```


# 17. Run a multivariable Cox model to determine the hazard rate of DEATH in current smokers vs. non-current smokers adjusted for age. a. Report the HR and 95% CI and b. interpret the results.
```{r}
Cox_mod_adj<-coxph(Surv(TIMEDTH, DEATH)~cursmoke + AGE, frm, ties="efron")
summary(Cox_mod_adj)

# HR = 1.57 (95% CI 1.42-1.75)
# Interpretation: After adjusting for age, current smoker respondents had a 1.57 (95% CI 1.42-1.75) times higher hazard of death than those who were not current smokers.

```

# 18. Determine if sex modifies the association between the age-adjusted hazard rate of DEATH in current smokers vs. non-current smokers. Report the results of your statistical test and interpret the results.
```{r}
Cox_mod_int<-coxph(Surv(TIMEDTH, DEATH)~cursmoke + AGE + SEX+ AGE*SEX, frm, ties="efron")
summary(Cox_mod_int)
```

```{r}
# Evaluate whether the interaction term is significant using the LR test
lrtest(Cox_mod_adj, Cox_mod_int)
#Interpretation: There is significant effect modification for the association between smoking status and death by sex.(p-value for lr test>0.05)
```


# 19. If sex modifies the association, run age-adjusted models to report associations stratified by sex. a. Report the age-adjusted HRs and 95% CIs for each sex and b. Interpret the results.
```{r}
# Male
Cox_mod_adj_male<-coxph(Surv(TIMEDTH, DEATH)~cursmoke + AGE, frm[which(frm$sex=="Male"),])
summary(Cox_mod_adj_male)
#HR(Male) = 1.53 (95% CI: 1.33-1.76)
# Interpretation: 

# Female
Cox_mod_adj_female<-coxph(Surv(TIMEDTH, DEATH)~cursmoke + AGE, frm[which(frm$sex=="Female"),])
summary(Cox_mod_adj_female)
# HR(Female) = 1.40 (95% CI: 1.19-1.65)
# Interpretation: 
```
# 20. Should the models be adjusted for sex or should sex-specific estimates be provided separately? Explain the rationale for your answer.
```{r}
#Sex-specific estimates should be provided separately because the stratum-specific HR differs by more than 10%. Therefore, sex should be considered as an effect modifier of the association between DEATH and current smoking status rather than as a confounder.
```


# Of note: the linearity assumption should be evaluated by age. However for the purpose of time, we have skipped that step in this exercise as well as other assumption tests.

# 21. Extra credit. Using the Cox model with the following right hand side: cursmoke + AGE + sex+ sex*cursmoke, determine the hazard rate of death in a 70 year old man who is a current smoker compared to a 80 year old man who is not a current smoker.  a. Report the HR and 95% CI. b. Interpret the results. c. How does this HR compare to the HR if they were both non-smokers? Hint you will need to calculate another HR for non-smokers.













